Impact of grids and dynamical cores in CESM2.2 on the meteorology and climate of the Arctic

### 2021-05-05 arcticgrids.tex
Updates from Overleaf

The meteorology and climate of the Arctic is characterized by a range of processes and scales that are difficult to represent in GCMs \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. For example, while synoptic scale storms are generally well represented at typical GCM resolutions of 1 to 2 degrees \cite{JW2006QJR,IPCCAR5}, mesoscale Polar Lows are not well resolved at these resolutions. These mesoscale systems are prevalent during the cold season and produce gale-force winds that can induce large heat and moisture fluxes through the underlying sea-ice/ocean interface. The Arctic also contains the Greenland Ice Sheet (hereafter denoted as \textit{GrIS}). While it blankets the largest island in the world (Greenland), many of the processes that control the GrIS annual surface mass balance (the integrated sum of precipitation and melting) are only partially resolved at typical GCM resolutions. For example, GrIS precipitation is typically confined to the ice-sheet margins (predominately the southeastern margin) where orographic precipitation is generated by steep topographic slopes. Moreover, GrIS ablation areas (marginal regions where seasonal melting exceeds the annual mass input from precipitation) are typically 10s to 100\,km wide and confined to low-level areas or regions with limited precipitation. GCMs struggle to resolve the magnitude and extent of these features \cite{VETAL2018TC}, which can lead to unrealistic ice sheet growth in models with an interactive ice sheet component \cite<e.g.,>{LETAL2020JAMES}.

@@

The atmospheric component of CESM2.2, the Community Atmosphere Model, version 6.3 (CAM; \url{https://ncar.github.io/CAM/doc/build/html/index.html}), supports a diverse number of atmospheric dynamical cores. These range from dycores using latitude-longitude grids, the finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, the spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM, and the least supported of all the dycores in the current model. FV3 is the newest dycore in CAM, but it was not fully incorporated at the time this work commenced, and so is omitted from this study. The authors instead focus on the two most well supported dycores in CAM, SE and FV.

\subsubsection{Finite-volume dynamical core}

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a spherical latitude-longitude grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to Eulerian reference grid in the vertical \cite{L2004MWR}, using a hybrid-pressure vertical coordinate. Hyperviscous damping is applied to the divergent modes while Laplacian damping is applied to momentum in the top few layers, referred to as a \textit{sponge layer} \cite{L2011IJHPC}. A polar filter is used to avoid computational instability due to the convergence of the meridians, allowing for a more practical time-step. It takes the form of a Fourier filter in the zonal direction, with the damping coefficients increasing monotonically poleward \cite{ST1995GEOS}.

\subsubsection{Spectral-element dynamical core}

The SE dycore is a hydrostatic model that integrates the equations of motion using a high-order continuous Galerkin method \cite{TTI1997JCP,DetAl2012IJHPCA}. The computational domain is a cubed-sphere grid tiled with quadrilateral elements (e.g., Figure~\ref{fig:vr-grids}). Each element contains a fourth order basis set in each horizontal direction, with the solution defined at the roots of the basis functions, the Gauss-Lobatto-Legendre (GLL) quadrature points. This results in 16 GLL nodal points within each element, with 12 of the points lying on the (shared) element boundary. Communication between elements happens via the direct stiffness summation \cite{canuto2007}, which applies a numerical flux to the element boundaries that reconciles overlapping nodal values and produces a continuous global basis set.

As with the FV dycore, the dynamics evolve in floating Lagrangian layers that are subsequently mapped to a Eulerian reference grid. A dry mass vertical coordinate was more recently implemented for thermodynamic consistency with condensates \cite{LetAl2018JAMES}. The 2D dynamics have no implicit dissipation and so hyperviscosity operators are applied to all prognostic variables to remove spurious numerical errors \cite{DetAl2012IJHPCA}. Laplacian damping is applied in the sponge layer.

@@

For spectral-element grids with quasi-uniform grid spacing, a variant in which tracer advection is computed using the Conservative Semi-Lagrangian Mulit-tracer tranport scheme is used instead \cite<CSLAM;>{LTOUNGK2017MWR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a seperate grid from the spectral-element dynamics, through dividing each element into $3\times3$ quasi-equal area control volumes. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the default SE dycore in which the physics are evaluated at the GLL nodal points \cite{HL2018MWR}.

@@

The CAM convention is that the physics package determines the vertical resolution. Since all runs use CAM6 physics, all grids and dycores use 32 levels in the vertical, with a model top of about $1 \ hPa$ or about $40 \ km$. The physics time-step, in contrast, is dependent on grid resolution. Increases in horizontal resolution permit faster vertical velocities that reduce characteristic time-scales, and so the physics time-step should be reduced to avoid large time truncation errors \cite{HR2018JAMES}. The $ARCTIC$ and $ARCTICGRIS$ grids are therefore run with a $4\times$ and $8\times$ reduction in physics time-step relative to the default 1800 s time-step used in coarser, uniform resolution grids.

@@

Before delving into the simulated characteristics of the Arctic, an understanding of the global mean differences between the grids and dycores are assessed, Figure~\ref{fig:dT-lores} shows the climatological, zonal mean height plots expressed as differences between the uniform resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes and throughout the depth of the troposphere. This is a common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}, and \cite{HK2020QJRMS} has shown that this occurs in CAM due to greater resolved vertical velocities that in turn, facilitate greater condensational heating in the macrophyiscs routine in CLUBB. The right columns in Figure~\ref{fig:dT-lores} supports this interpretation, which shows an increase in the climatological CLUBB heating in the mid-latitudes in the $f09$ grid.

As the SE dycore is less diffusive than the FV dycore, the resolved vertical velocities are larger in the SE dycore, and so a modest, resolution-like sensitivity occurs in which $ne30pg3$ is warmer than $f09$ (Figure~\ref{fig:dT-lores}). The stratosphere has a different response, in which $ne30pg3$ is much cooler than $f09$ in the mid-to-high latitudes. The differences in temperature between $ne30pg3$ and $ne30pg2$ are small, with a slight warming near the tropopause at high latitudes. This is consistent with the similar climates found between these grids in \cite{HETAL2019JAMES}.

Comparing the variable-resolution grids to the uniform resolution grids is complicated because we are simultaneously increase the grid resolution and reducing the physics time-step, both which noticeably impact the solution \cite{W2008TELLUS}. An additional $ne30pg3$ simulations is run with the physics time-step used in the $ARCTIC$ grid, referred to as $ne30pg3^{*}$. Figure~\ref{fig:dT-hires} shows the change in climatological temperature in zonal-mean height space between $ne30pg3^{*}$ and $ne30pg3$. A similar warming response to increasing resolution occurs when the time-step is reduced, and the mechanism is similar in that the shorter time-step facilitates greater condensational heating by CLUBB. Figure~\ref{fig:dT-hires} shows the difference in climatological temperature between the $ARCTIC$ grid and the $ne30pg3^{*}$ grid. The greater condensational heating and warmer temperatures are confined to the regionally refined region when the impact of physics time-steps is removed from the analysis.

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_dhgt_panel_STEND_CLUBB-hires.pdf}
\end{center}
\caption{.}
\label{fig:dT-hires}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_tseries_GRIS.pdf}
\end{center}
\caption{.}
\label{fig:tseries}
\end{figure}

>>>>

The meteorology and climate of the Arctic is characterized by a range of processes and scales that are difficult to represent in GCMs \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. For example, while synoptic scale storms are generally well represented at typical GCM resolutions of 1 to 2 degrees (REF: 2013 IPCC report, 9.1.3.3, Jablonowski and Williamson), mesoscale Polar Lows are not well resolved at these resolutions. These mesoscale systems are prevalent during the cold season and produce gale-force winds that can induce large heat and moisture fluxes through the underlying sea-ice/ocean interface. The Arctic also contains the Greenland Ice Sheet (hereafter denoted as \textit{GrIS}). While it blankets the largest island in the world (Greenland), many of the processes that control the GrIS annual surface mass balance (the integrated sum of precipitation and melting) are only partially resolved at typical GCM resolutions. For example, GrIS precipitation is typically confined to the ice-sheet margins (predominately the southeastern margin) where orographic precipitation is generated by steep topographic slopes. Moreover, GrIS ablation areas (marginal regions where seasonal melting exceeds the annual mass input from precipitation) are typically 10s to 100\,km wide and confined to low-level areas or regions with limited precipitation. GCMs struggle to resolve the magnitude and extent of these features \cite{VETAL2018TC}, which can lead to unrealistic ice sheet growth in models with an interactive ice sheet component \cite[e.g.,][]{Lofverstrom_EA_2020_JAMES}.

@@

The atmospheric component of CESM2.2, the Community Atmosphere Model, version 6.3 (CAM; \url{https://ncar.github.io/CAM/doc/build/html/index.html}), supports a number of different atmospheric dynamical cores. These include dycores using latitude-longitude grids, such as finite-volume \cite<FV;>{L2004MWR} and eulerian spectral transform \cite<EUL;>{CETAL006JC} models, and dycores built on unstructured grids, including spectral-element \cite<SE;>{LetAl2018JAMES} and finite-volume 3 \cite<FV3;>{PL2007JCP} models. The EUL dycore is the oldest dycore in CAM and the least supported of all the dycores in the current model. FV3 is the newest dycore in CAM, but it was not fully incorporated at the time this work commenced; both the EUL and FV3 dycores are omitted from this study. As such, the results presented in this study are comparing the performance of the SE and FV dycores.

\subsubsection{Finite-volume (FV) dynamical core}

The FV dycore is a hydrostatic model that integrates the equations of motion using a finite-volume discretization on a spherical latitude-longitude grid \cite{LR1997QJR}. The 2D dynamics evolve in floating Lagrangian layers that are periodically mapped to Eulerian reference grid in the vertical \cite{L2004MWR}, using a hybrid-pressure vertical coordinate. Hyperviscous damping is applied to the divergent modes while Laplacian damping is applied to momentum in the top few layers, referred to as a \textit{sponge layer} \cite{L2011IJHPC}. A zonal Fourier filter --- with damping coefficients increasing monotonically in the poleward direction \cite{ST1995GEOS} --- is applied to suppress numerical instabilities due to grid convergence at the North and South Pole. This so called \textit{polar filer} allows for a longer time-step, but it has the disadvantage that atmospheric fields are artificially smoothed at high latitudes.

\subsubsection{Spectral-element (SE) dynamical core}

The SE dycore is a hydrostatic model that integrates the equations of motion using a high-order continuous Galerkin method \cite{TTI1997JCP,DetAl2012IJHPCA}. The computational domain is a cubed-sphere grid tiled with quadrilateral elements (e.g., Figure~\ref{fig:vr-grids}). Each element contains a fourth order basis set in the two horizontal directions, with the solution defined at the roots of the basis functions, the Gauss-Lobatto-Lagendre (GLL) quadrature points. This results in 16 GLL nodal points within each element, with 12 of the points lying on the (shared) element boundary. Communication between elements happens via the direct stiffness summation \cite{canuto2007}, which applies a numerical flux to the element boundaries that reconciles overlapping nodal point values and produces a continuous global basis set.

As with the FV dycore, the dynamics evolve in floating Lagrangian layers that are subsequently mapped to an Eulerian reference grid. A dry mass vertical coordinate was recently implemented for thermodynamic consistency with condensates \cite{LetAl2018JAMES}. The 2D dynamics have no implicit dissipation and so hyperviscosity operators are applied to all prognostic variables to remove spurious numerical errors \cite{DetAl2012IJHPCA}. Laplacian damping is applied in the sponge layer.

@@

For spectral-element grids with quasi-uniform grid spacing, a variant in which tracer advection is computed using the Conservative Semi-Lagrangian Multi-tracer transport scheme (CSLAM) is used instead \cite<CSLAM;>{LTOUNGK2017MWR}. CSLAM has improved tracer property preservation and accelerated multi-tracer transport. It uses a separate grid from the spectral-element dynamics, through dividing each element into $3\times3$ control volumes with quasi-equal area. The physical parameterizations are computed from the state on the CSLAM grid, which has clear advantages over the default SE dycore in which the physics are evaluated at the GLL nodal points \cite{HL2018MWR}.

@@

@@

Before delving into the simulated characteristics of the Arctic, an understanding of the global mean differences between the grids and dycores are assessed, Figure~\ref{fig:dT-lores} shows the climatological, zonal mean height plots expressed as differences between the uniform resolution grids and dycores. The $f09$ grid is warmer than the $f19$ grid, primarily in the mid-to-high latitudes and throughout the depth of the troposphere.This is common response to increasing horizontal resolution in GCMs \cite{PS2002CD,RETAL2006JC}, and \cite{HK2020QJRMS} has shown that this occurs in CAM due to greater resolved vertical velocities that in turn, facilitate greater condensational heating in the macrophyiscs routine in CLUBB. The right columns in Figure~\ref{fig:dT-lores} supports this interpretation, which shows an increase in the climatological CLUBB heating in the mid-latitudes in the $f09$ grid.

As the SE dycore is less diffusive than the FV dycore, the resolved vertical velocities are larger in the SE dycore, and so a modest, resolution-like sensitivity occurs in which $ne30pg3$ is warmer than $f09$ (Figure~\ref{fig:dT-lores}). The stratosphere has a different signal, in which $ne30pg3$ is much cooler than $f09$ in the mid-to-high latitudes. The differences in temperature between $ne30pg3$ and $ne30pg2$ are small, with a slight warming near the tropopause at high latitudes. This is consistent with the similar climates using these grids shown in \cite{HETAL2019JAMES}.

In comparing the variable-resolution grids to the uniform resolution grids is convoluted because not only is the resolution higher in refinement region, but the physics time-step is shorter in the variable-resolution runs. An additional $ne30pg3$ simulations was run with physics time-step used in the $ARCTIC$ grid, referred to as $ne30pg3^{*}$. Figure~\ref{fig:dT-hires} shows the changes in climatological temperature in zonal-mean height space between $ne30pg3^{*}$ and $ne30pg3$. A similar warming response to increase resolution occurs when the time-step is reduced, and the mechanism is similar in that this faciliates greater condensational heating by CLUBB. Figure~\ref{fig:dT-hires} shows difference in climatolgical temperature between the $ARCTIC$ grid and the $ne30pg3^{*}$ grid. The greater condensational heating and warmer temperatures are confined to the regionally refined region when the impact of physics time-steps is removed from the analysis.

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_dhgt_panel_STEND_CLUBB_phytstep_JJA.pdf}
\end{center}
\caption{.}
\label{fig:dT-hires}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_tseries_GRIS_map_TO_f19.pdf}
\end{center}
\caption{.}
\label{fig:tseries}
\end{figure}
